---
title: "CO2 Emissions Inspection"
output: html_notebook
---

# Import the Dataset

# (Change the path if you want to run!!)

```{r}
# Load the dataset 
df <- read.csv('/Users/berdn90s/Downloads/CO2 Emissions Dataset/visualizing_global_co2_data.csv', stringsAsFactors = FALSE)

# Overwrite df with only the selected columns
df <- df[, c(
  "country",
  "year",
  "population",
  "gdp",
  "co2",
  "co2_per_capita",
  "co2_per_gdp",
  "coal_co2",
  "oil_co2",
  "gas_co2",
  "cement_co2",
  "cumulative_co2",
  "share_global_co2",
  "temperature_change_from_co2"
)]

# View the structure of the updated dataset
str(df)
```

## Check Missing Values

```{r}
# Calculate the number of NA values per column
na_counts <- colSums(is.na(df))

# Calculate the percentage of NA values per column
na_percentage <- (na_counts / nrow(df)) * 100

# Combine the counts and percentages into a data frame
na_summary <- data.frame(
  Column = names(na_counts),
  NA_Count = na_counts,
  NA_Percentage = na_percentage
)

# Display the NA summary
print(na_summary)
```

## Handling Missing Values

```{r}
# Fill missing values with median for numerical columns
df$co2[is.na(df$co2)] <- median(df$co2, na.rm = TRUE)
df$co2_per_capita[is.na(df$co2_per_capita)] <- median(df$co2_per_capita, na.rm = TRUE)
df$cumulative_co2[is.na(df$cumulative_co2)] <- median(df$cumulative_co2, na.rm = TRUE)
df$share_global_co2[is.na(df$share_global_co2)] <- median(df$share_global_co2, na.rm = TRUE)
df$temperature_change_from_co2[is.na(df$temperature_change_from_co2)] <- mean(df$temperature_change_from_co2, na.rm = TRUE)
df$population[is.na(df$population)] <- median(df$population, na.rm = TRUE)
```

## Drop Columns with High Missing Values

```{r}
# Drop columns with >50% missing values
df <- df[, !names(df) %in% c("gdp", "co2_per_gdp", "coal_co2", "oil_co2", "gas_co2", "cement_co2")]

# Preview the cleaned dataset
head(df)
```

## Descriptive Statistics

```{r}
# Summary statistics for the dataset
summary(df)

# Frequency table for 'country' column
country_freq <- table(df$country)
print(country_freq)
```

# Filter Data for Europe and Asia

```{r}
# Create a vector of European and Asian countries
europe_asia_countries <- c(
  "Afghanistan", "Albania", "Armenia", "Austria", "Azerbaijan", "Bangladesh", 
  "Belgium", "Bhutan", "Bosnia and Herzegovina", "Bulgaria", "China", "Croatia",
  "Cyprus", "Czech Republic", "Denmark", "Estonia", "Finland", "France",
  "Georgia", "Germany", "Greece", "Hungary", "India", "Indonesia", "Iran",
  "Iraq", "Ireland", "Israel", "Italy", "Japan", "Jordan", "Kazakhstan",
  "Kuwait", "Kyrgyzstan", "Laos", "Latvia", "Lebanon", "Lithuania", 
  "Luxembourg", "Malaysia", "Maldives", "Moldova", "Mongolia", "Montenegro", 
  "Myanmar", "Nepal", "Netherlands", "North Korea", "Norway", "Oman", "Pakistan",
  "Palestine", "Philippines", "Poland", "Portugal", "Qatar", "Romania", 
  "Russia", "Saudi Arabia", "Serbia", "Singapore", "Slovakia", "Slovenia", 
  "South Korea", "Spain", "Sri Lanka", "Sweden", "Switzerland", "Syria",
  "Tajikistan", "Thailand", "Turkey", "Turkmenistan", "Ukraine", 
  "United Arab Emirates", "Uzbekistan", "Vietnam", "Yemen"
)

# Subset the data to include only the specified countries
df_europe_asia <- df %>% filter(country %in% europe_asia_countries)

# Get the unique countries in the Europe and Asia dataset
unique_countries_europe_asia <- unique(df_europe_asia$country)

# Print the unique countries
print(unique_countries_europe_asia)
```

# Visualizations

### 1. Top 10 CO2-Emitting Countries in Europe and Asia

```{r}
# Summarize average CO2 emissions for each country
top_countries_eu_asia <- df_europe_asia %>%
  group_by(country) %>%
  summarise(avg_co2 = mean(co2, na.rm = TRUE)) %>%
  arrange(desc(avg_co2)) %>%
  slice_head(n = 10)

# Bar chart
ggplot(top_countries_eu_asia, aes(x = reorder(country, -avg_co2), y = avg_co2, fill = country)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Top 10 CO2-Emitting Countries in Europe and Asia",
    x = "Country",
    y = "Average CO2 Emissions"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 2. CO2 Emissions Over Time

```{r}
# Line plot for CO2 emissions over time for Europe and Asia
ggplot(df_europe_asia, aes(x = year, y = co2, group = country, color = country)) +
  geom_line() +
  labs(
    title = "CO2 Emissions Over Time in Europe and Asia",
    x = "Year",
    y = "CO2 Emissions"
  ) +
  theme_minimal() +
  theme(legend.title = element_text(size = 10), legend.text = element_text(size = 8))
```

### 3. Population vs CO2 Emissions

```{r}
# Scatter plot for population vs CO2 emissions
ggplot(df_europe_asia, aes(x = population, y = co2)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Population vs CO2 Emissions in Europe and Asia",
    x = "Population",
    y = "CO2 Emissions"
  ) +
  theme_minimal()
# The red line represents the linear regression fit.
```

### 4. Distribution of CO2 Emissions

```{r}
# Histogram of CO2 emissions in Europe and Asia
ggplot(df_europe_asia, aes(x = co2)) +
  geom_histogram(binwidth = 50, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of CO2 Emissions in Europe and Asia",
    x = "CO2 Emissions",
    y = "Frequency"
  ) +
  theme_minimal()
```

### 5. Frequency of CO2 Emission Categories

```{r}
# Adjust thresholds for CO2 emissions bins
df_europe_asia$CO2_Emissions_Binned <- cut(
  df_europe_asia$co2,
  breaks = c(0, 100, 1000, 5000, 15000),
  labels = c("Very Low (<100)", "Low (100-1000)", "Moderate (1000-5000)", "High (>5000)")
)

# Count the frequency of each category
freq_table <- df_europe_asia %>%
  group_by(CO2_Emissions_Binned) %>%
  summarise(Count = n())

# Bar chart with new thresholds
ggplot(freq_table, aes(x = CO2_Emissions_Binned, y = Count, fill = CO2_Emissions_Binned)) +
  geom_bar(stat = "identity", color = "black") +
  labs(
    title = "Frequency of CO2 Emission Categories in Europe and Asia (Adjusted Thresholds)",
    x = "CO2 Emission Categories",
    y = "Number of Countries"
  ) +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

# Correlation Inspection

### Correlation Matrix

```{r}
# Compute the correlation matrix
numeric_cols <- df_europe_asia %>%
  select_if(is.numeric)

cor_matrix <- cor(numeric_cols, use = "complete.obs")

# Print the correlation matrix
print(cor_matrix)
```

### Correlation Heatmap

```{r}
library(reshape2)

# Convert the correlation matrix to long format for ggplot2
cor_long <- melt(cor_matrix)

# Create a heatmap
ggplot(cor_long, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue", high = "red", mid = "white", 
    midpoint = 0, limit = c(-1, 1), space = "Lab"
  ) +
  labs(
    title = "Correlation Heatmap",
    x = "Variables",
    y = "Variables"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

### Insights from the Correlation Heatmap

-   **Population and CO2**: There is a strong correlation between population and CO2 emissions, which we will explore further.
-   **Other Strong Correlations**: Variables like `share_global_co2` and `co2` are strongly correlated, but this is expected and not insightful for further analysis.

# Clustering Analysis

### K-Means Clustering

```{r}
# Select numeric columns for clustering
numeric_cols <- df_europe_asia %>%
  select(co2, population, co2_per_capita)

# Scale the data for clustering
scaled_data <- scale(numeric_cols)

# Perform k-means clustering (e.g., 3 clusters)
set.seed(123)
kmeans_result <- kmeans(scaled_data, centers = 3, nstart = 25)

# Add cluster assignments to the dataset
df_europe_asia$Cluster <- as.factor(kmeans_result$cluster)

# Visualize clusters (Example: CO2 vs Population)
ggplot(df_europe_asia, aes(x = population, y = co2, color = Cluster)) +
  geom_point() +
  labs(
    title = "K-Means Clustering: CO2 vs Population",
    x = "Population",
    y = "CO2 Emissions"
  ) +
  theme_minimal()
```

### Interpretation of K-Means Clustering

-   **Cluster 1 (Red)**: Countries with high population and high CO2 emissions, likely industrialized nations with significant environmental impact.
-   **Cluster 2 (Green)**: Countries with moderate populations and CO2 emissions, potentially developing or medium-sized nations.
-   **Cluster 3 (Blue)**: Countries with very low populations and CO2 emissions, likely smaller or less industrialized regions.

This clustering highlights the strong correlation between population size and CO2 emissions.

# Regression Model

### Spline Regression: Predicting CO2 Emissions

```{r}
# Fit a spline model
spline_model <- lm(co2 ~ bs(population, degree = 3, knots = c(5e+08, 1e+09)), data = df_europe_asia)

# Summary of the spline model
summary(spline_model)

# Predict and visualize the spline fit
df_europe_asia$predicted_co2_spline <- predict(spline_model, newdata = df_europe_asia)

ggplot(df_europe_asia, aes(x = population, y = co2)) +
  geom_point(color = "blue", alpha = 0.5) +
  geom_line(aes(y = predicted_co2_spline), color = "red") +
  labs(
    title = "Spline Regression: CO2 vs Population",
    x = "Population",
    y = "CO2 Emissions"
  ) +
  theme_minimal()
```

### Interpretation of Spline Regression

-   **Spline Regression** divides the data into segments using knots (breakpoints) and fits separate polynomial functions to each segment, ensuring smooth transitions.
-   **Knots**: Split the population into meaningful ranges (e.g., at 500 million and 1 billion).
-   **R-squared = 0.7681**: The model explains **76.8%** of the variability in `co2` emissions based on `population`.
-   This approach captures the nonlinear relationship better than linear models, providing a significant improvement.
